\begin{center} \bf \sffamily \LARGE Prisoner's dilemma with stochastic payoffs\\[-1cm] \end{center}

### 2021-04-22 main.tex
about to move things around in introduction

@@

@@

Evolutionary game theory is a mathematical framework for modeling evolution in
biological, social and economical systems. %ToDo references
In evolutionary models, the success of an individual is not constant but it
depends on the composition of the population. Individuals interact according to
the rules of a pre defined game and they payoffs they achieved are translated
into fitness.

Evolutionary models often work with idealized assumptions when estimating the
fitness of a individual. It is commonly assumed that an individual will interact
several times and with all other players in the population. Equivalently their
fitness is given by the mean payoff they achieved over the different types in
the population after multiple intercalations. These payoff is referred to as the
\textit{expected} payoff.

Initially this assumption implies that selection occurs much more slowly than
the interaction between individuals. However, experimental studies show that
this may not always be the case in biology %ToDo references.
In~\cite{Roca2006} this assumption is realized and it shows that rapid selection
affects evolutionary dynamics in such a dramatic way that for some games it even
changes the stability of equilibria.

Moreover, estimating expected payoffs assume that individuals have a perfect a
memory. Their recall several of their interactions with each player in the
population. This creates a curious inconsistency. With a only few notable
exceptions~\cite{Hauert1997, Stewart2016}, evolutionary models focus on naive
subjects who can only choose from a restricted set of
strategies~\cite{Nowak1992tit}, or who do not remember anything beyond the
outcome of the very last round~\cite{Baek2016} when modeling how individuals
make decisions in each round. However, when modeling how individuals update
their strategies over time, individuals are assumed to have perfect memory.

We will explore how robust our understanding of cooperation is as models deviate
from the perfect memory assumption. We start by considering two extreme
scenarios. The first is the classical scenario of the expected payoffs and the
alternative scenario where individuals update their strategies only based on the
very last payoff they obtained. For our second approach we allow individuals to
use more memory. We refer to these as the \textit{stochastic} payoffs of
individuals.

Early results for the repeated prisoner’s dilemma and for the snowdrift game
suggest that memory size can have a considerable effect on the evolutionary
dynamics. As a rule of thumb, we observe that individuals with limited memory
tend to adopt less generous strategies and they achieve less cooperation than in
the classical scenario. In contrast, in the stag-hunt and the harmony game, the
impact of memory is less striking. Here individuals with limited memory perform
nearly as well as individuals with full memory.

@@

Evolutionary game dynamics have extensively been studied using stochastic
individual-based models that make use of different rules to determine how
successful strategies spread. There are two classes have been used extensively
based on~\cite{Wu2015} (i) \textit{fitness-based processes} in which an
individual chosen proportional to fitness reproduces and the offspring replaces
a randomly chosen individual (ii) \textit{pairwise comparison} processes in
which a pair of individuals is chosen, and where subsequently one of these
individuals may adopt the strategy of the other. This adoption occurs with a
probability that depends on the fitness of both individuals, such that better
players are more likely to be imitated than those who do worse.

We us a pairwise comparison processes with a population of $N$ individuals. In
our model mutation happens rarely and such each individual can be of one of the
two types, either of the \textit{resident} or either of the \textit{mutant}.

The fitness of the individuals will be based on the payoffs they receive from
their interactions. he individuals of our population do not interact for a given
number of turns, on the contrast, we consider that each match has a probability
$\delta$ to continue following each turn. In each turn individuals can decide whether
to cooperate (C) or to defect (D). The
payoffs in a given round depend on the player's and co players decisions. We
consider all the important case of symmetric \(2 \times 2\) games, in which the
payoffs are given by the following matrix:

\begin{equation}\label{eq:game}
    \begin{pmatrix}
        R & S  \\
        T & P
    \end{pmatrix}
\end{equation}

These cases include the Snowdrift~\cite{sugden2004economics} ($T>R>P>S$), Stag
Hunt~\cite{skyrms2001stag} ($T>R>P>S$), Harmony~\cite{Licht1999} ($T>R>P>S$) and
the Prisoner's Dilemma ($T>R>P>S$) games. The results based on these well know
\(2 \times 2\) games are discussed in Section~\ref{section:2_by_2_games}.

@@

The relative influence of the payoffs on the adopt the strategy of the other is
controlled by an external parameter, the so called intensity of selection
$\beta$.

The space of strategies in repeated is infinite. To continue with our
evolutionary analysis, we assume herein that individuals at most make use of
simple \textit{reactive strategies}. Reactive strategy are a set of memory-one
strategies that only take into account the previous action of the opponent.
Reactive strategies can be written explicitly as a vector \(\in \R_{3}\). More
specifically a reactive strategy \(s\) is given by \(s=(y, p, q)\) where \(y\)
is the probability that the strategy opens with a cooperation and \(p, q\) are
the probabilities that the strategy cooperates given that the opponent
cooperated and defected equivalently.

@@

It can shown that, \(!(1\!-\!\delta)\mathbf{v_0}(I_4-\delta M)^{-1}\) and with
some algebraic manipulation we derive to Equation~\ref{Eq:LastRound}.

@@

>>>>

\usepackage{tikz}
\usetikzlibrary{arrows}

\tikzset{
  treenode/.style = {align=center, inner sep=0pt, text centered,
    font=\sffamily},
  arn_n/.style = {treenode, circle, white, font=\sffamily\bfseries, draw=black, inner sep=-6pt,
    fill=black, text width=1.5em},% arbre rouge noir, noeud noir
  arn_r/.style = {treenode, circle, red, fill=red, 
    text width=1.5em, very thick, inner sep=-6pt},% arbre rouge noir, noeud rouge
  arn_x/.style = {treenode, rectangle, draw=black,
    minimum width=0.5em, minimum height=0.5em}% arbre rouge noir, nil
}

@@

\begin{abstract}

\end{abstract}

@@

Evolutionary game theory is a mathematical framework for modeling evolution in
biological, social and economical systems. %ToDo references
Compared to classical game theory where it is assumed that players are rational,
evolutionary game theory does not require individuals to be rational. Instead
individuals adapt their strategies based on exploration and mutation. In this
context, strategies spread based on fitness either because the individuals who
adopt them have more offspring (fitness-based processes), or they are imitated
more often (pairwise comparison processes)~\cite{Wu2015}. Fitness is not
constant but it depends on the composition of the population. Individuals
interact with members of the population according to pre-define games, and their
payoffs are then translated into fitness.

Evolutionary models often work with idealized assumptions when estimating
fitness. It is commonly assumed that fitness is given by the mean payoff an
individual achieved over the different types in the population following
multiple intercalations. Referred to as the \textit{expected payoffs}. Expected
payoffs imply that individuals have a perfect a memory, as they are able to
recall several of their interactions with each player in the population. However
most evolutionary models, with a few notable exceptions~\cite{Hauert1997,
Stewart2016}, focus on naive subjects who can only choose from a restricted set
of strategies~\cite{Nowak1992tit}, or who do not remember anything beyond the
outcome of the very last round~\cite{Baek2016} when modeling how individuals
make decisions in each round. This creates a curious inconsistency. Individuals
make decisions in each round whist having very limited memory, however, when
modeling how individuals update their strategies over time, individuals are
assumed to have perfect memory.

One of the most important applications of evolutionary game theory is the
evolution of cooperation. Though cooperation can been seen to be at odd, several
mechanisms can allow for cooperation to emerge. One of these mechanisms is
direct reciprocity. However, given these idealized assumptions it feels like our
understanding of reciprocity is thus still incomplete. In this work we will
explore how robust our understanding of cooperation is as models deviate from
the perfect memory assumption.

This work is not the first to relax assumption made at the fitness level.
In~\cite{Roca2006} relax the assumption that selection occurs much more slowly
than the interaction between individuals. They results shows that rapid
selection affects evolutionary dynamics in such a dramatic way that for some
games it even changes the stability of equilibria.

We start by considering two extreme scenarios. The first is the classical
scenario of the expected payoffs and the alternative scenario where individuals
update their strategies only based on the very last payoff they obtained. We
refer to these as the \textit{stochastic} payoffs of individuals. In
Section~\ref{section:donation}, we explore the effect of the payoffs in the
cooperation rate given the donation game. We explore the difference between the
ratios as well as the effect of different model parameters. In
Section~\ref{section:2_by_2_games} we explore the cooperation rate for all the
important case of symmetric \(2 \times 2\) games. For the later part of the
section we also allow individuals to use more memory.

% Early results for the repeated prisoner’s dilemma and for the snowdrift game
% suggest that memory size can have a considerable effect on the evolutionary
% dynamics. As a rule of thumb, we observe that individuals with limited memory
% tend to adopt less generous strategies and they achieve less cooperation than in
% the classical scenario. In contrast, in the stag-hunt and the harmony game, the
% impact of memory is less striking. Here individuals with limited memory perform
% nearly as well as individuals with full memory.

@@

In order to account for the difference in the robustness of cooperation among
individuals based on their payoff memory, we consider a population of \(N\)
players, where \(N\) is even, and where mutations are sufficiently rare. At any
point in time the there are at most two different strategies are present in the
population. A \textit{resident} strategy and a \textit{mutant} strategy. Suppose
there are \(N - k\) players who use the resident strategy whereas \(k\) players
use the mutant strategy. Each step of the evolutionary process consists of two
stages, a game stage and an updating stage.

1. In the game stage, each player is randomly matched with some other player in
the population to interact in a number of instances of the game
\(\begin{pmatrix} R & S  \\ T & P \end{pmatrix}\). The number of interactions is
not fixed, on the contrast we consider that another interaction has a
probability $\delta$ to continue following each turn. We assume herein that
individuals at most make use of simple \textit{reactive strategies} make
decisions in each round. Reactive strategy are a set of memory-one strategies
that only take into account the previous action of the opponent. Reactive
strategies can be written explicitly as a vector \(\in \R_{3}\). More
specifically a reactive strategy \(s\) is given by \(s=(y, p, q)\) where \(y\)
is the probability that the strategy opens with a cooperation and \(p, q\) are
the probabilities that the strategy cooperates given that the opponent
cooperated and defected equivalently.

2. In the updating stage, two players are randomly drawn from the population, a
`learner' and a `role model'. Given that the learner's payoff $u_L\!\in\!
\mathcal{U}$ and that the role model's payoff $u_{RL}\!\in\! \mathcal{U}$, we
assume the learner adopts the role model's strategy with probability

\begin{equation} \label{Eq:rho}
\rho(u_L, u_{RL}) = \frac{1}{1\!+\!\exp\big[ \!-\!\beta (u_E\!-\!u_{RL}) \big]}.
\end{equation}

where $\beta\!\ge\!0$ corresponds  to relative influence of the payoffs on the
adopt the strategy of the other is controlled by an external parameter, the so
called intensity of selection.

We iterate this basic evolutionary step until either the mutant strategy goes
extinct, or until it fixes in the population (in which case the mutant strategy
becomes the new resident strategy). After either outcome, we introduce a new
mutant strategy $S'_2\!=\!(y_2',p_2',q_2')$ (uniformly chosen from all reactive
strategies at random), and we set the number of mutants to $k\!=\!1$. This
process of mutation and fixation/extinction is then iterated many times.xw The
fixation probability of the mutant strategy then takes the standard form
\cite{nowak2004emergence},

\begin{equation}
\varphi = \frac{1}{1+\sum_{i=1}^{N-1}\prod_k^i \frac{\lambda^-_k}{\lambda^+_k}}.
\end{equation}

where \(\lambda^-_k, \lambda^+_k\) are the probabilities that the number of
mutants decreases and increases respectively.

We compare this process for stochastic payoff evaluation with the analogous
process where players update their strategies with respect to their {\it
expected} payoffs,

\begin{equation} \label{Eq:ExpPay}
  \begin{array}{lcrcr}
  \displaystyle \pi_1	&=	&\displaystyle \frac{N\!-\!k\!-\!1}{N-1}\cdot \langle\mathbf{v}(S_1,S_1),\mathbf{u}\rangle	&+	&\displaystyle\frac{k}{N-1}\cdot \langle\mathbf{v}(S_1,S_2),\mathbf{u}\rangle,\\[0.5cm]
  \displaystyle \pi_2	&=	&\displaystyle\frac{N-k}{N-1}\cdot \langle\mathbf{v}(S_2,S_1),\mathbf{u}\rangle&+	&\displaystyle\frac{k-1}{N-1}\cdot \langle\mathbf{v}(S_2,S_2),\mathbf{u}\rangle.\\
  \end{array}
\end{equation}

% \subsection{Fixation probabilities under stochastic payoff evaluation}
% Given that $N\!-\!k$ players use the resident strategy $S_1\!=\!(y_1,p_1,q_1)$ and that the remaining $k$ players use the mutant strategy $S_2\!=\!(y_2,p_2,q_2)$, the probability that the number of mutants increases by one in one step of the evolutionary process can be written as
% \begin{equation}
% \lambda^+_k=\frac{N\!-\!k}{N}\cdot \frac{k}{N}\cdot \sum_{u_1,u_2\in\mathcal{U}} x(u_1,u_2)\cdot \rho(u_1,u_2).
% \end{equation}
% In this expression, $(N\!-\!k)/N$ is the probability that the randomly chosen learner is a resident, and $k/N$ is the probability that the role model is a mutant. The sum corresponds to the total probability that the learner adopts the role model's strategy over all possible payoffs $u_1$ and $u_2$ that the two player may have received in their respective last rounds. We use $x(u_1,u_2)$ to denote the probability that the randomly chosen resident obtained a payoff of $u_1$ in the last round of his respective game, and that the mutant obtained a payoff of $u_2$. Given that the payoffs are $u_1$ and $u_2$, the imitation probability is then given by $\rho(u_1,u_2)$, as specified by Eq.~(\ref{Eq:rho}). The probability that the respective payoffs of the players are given by $u_1$ and $u_2$ can be calculated as
% \begin{equation}
% \setlength{\arraycolsep}{1pt}
% \begin{array}{llrl}
% x(u_1,u_2)	 &=&\displaystyle \frac{1}{N\!-\!1}\cdot  &v_{u_1}(S_1,S_2)\cdot 1_{(u_1,u_2)\in \mathcal{U}^2_F}\\[0.5cm]
% &+	
% &\displaystyle \left(1\!-\!\frac{1}{N\!-\!1}\right)  
% &\left[ \frac{k\!-\!1}{N\!-\!2}\frac{k\!-\!2}{N\!-\!3} v_{u_1}(S_1,S_2) v_{u_2}(S_2,S_2) + 
%  \frac{k\!-\!1}{N\!-\!2}\frac{N\!-\!k\!-\!1}{N\!-\!3} v_{u_1}(S_1,S_2) v_{u_2}(S_2,S_1)\right.\\[0.5cm]
% &&&\left. +\frac{N\!-\!k\!-\!1}{N\!-\!2}\frac{k\!-\!1}{N\!-\!3} v_{u_1}(S_1,S_1) v_{u_2}(S_2,S_2) + 
%  \frac{N\!-\!k\!-\!1}{N\!-\!2}\frac{N\!-\!k\!-\!2}{N\!-\!3} v_{u_1}(S_1,S_1) v_{u_2}(S_2,S_1)\right].
% \end{array}
% \end{equation}
%  The first term on the right side corresponds to the case that the learner and
%  the role model happened to be matched during the game stage, which happens
%  with probability $1/(N\!-\!1)$. In that case, we note that only those payoff
%  pairs can occur that are feasible in a direct interaction, $(u_1,u_2)\in
%  \mathcal{U}^2_F:=\big\{ (R,R), (S,T), (T,S), (P,P) \big\}$, as represented by
%  the respective indicator function. Otherwise, if the learner and the role
%  model did not interact directly, we need to distinguish four different cases,
%  depending on whether the learner was matched with a resident or a mutant, and
%  depending on whether the role model was matched with a resident or a mutant.

% \begin{equation}\label{eq:game}
%     \begin{pmatrix}
%         R & S  \\
%         T & P
%     \end{pmatrix}
% \end{equation}

% These cases include the Snowdrift~\cite{sugden2004economics} ($T>R>P>S$), Stag
% Hunt~\cite{skyrms2001stag} ($T>R>P>S$), Harmony~\cite{Licht1999} ($T>R>P>S$) and
% the Prisoner's Dilemma ($T>R>P>S$) games. The results based on these well know
% \(2 \times 2\) games are discussed in Section~\ref{section:2_by_2_games}.

@@

@@

It can shown that, \((1\!-\!\delta)\mathbf{v_0}(I_4-\delta M)^{-1}\) and with
some algebraic manipulation we derive to Equation~\ref{Eq:LastRound}.

@@

Equation~\ref{Eq:LastRound} is the probability vector that players \(S_1, S_2\)
are in each of the possible states of the last round. Given the population \(N\)
with \(k\) mutants in the last round there only possible combinations of interactions
are:

\begin{figure}
    \centering
    \includegraphics[width=.65\textwidth]{static/last_rounds_matches.tex}
\end{figure}

We compare this process for stochastic payoff evaluation with the analogous process where players update their strategies with respect to their {\it expected} payoffs,
\begin{equation} \label{Eq:ExpPay}
\begin{array}{lcrcr}
\displaystyle \pi_1	&=	&\displaystyle \frac{N\!-\!k\!-\!1}{N-1}\cdot \langle\mathbf{v}(S_1,S_1),\mathbf{u}\rangle	&+	&\displaystyle\frac{k}{N-1}\cdot \langle\mathbf{v}(S_1,S_2),\mathbf{u}\rangle,\\[0.5cm]
\displaystyle \pi_2	&=	&\displaystyle\frac{N-k}{N-1}\cdot \langle\mathbf{v}(S_2,S_1),\mathbf{u}\rangle&+	&\displaystyle\frac{k-1}{N-1}\cdot \langle\mathbf{v}(S_2,S_2),\mathbf{u}\rangle.\\
\end{array}
\end{equation}
In the limit of no discounting, $\delta\!\rightarrow\! 1$, this process based on expected payoffs has been considered in~\cite{imhof2010stochastic}.

\subsection{Fixation probabilities under stochastic payoff evaluation}
Given that $N\!-\!k$ players use the resident strategy $S_1\!=\!(y_1,p_1,q_1)$ and that the remaining $k$ players use the mutant strategy $S_2\!=\!(y_2,p_2,q_2)$, the probability that the number of mutants increases by one in one step of the evolutionary process can be written as
\begin{equation}
\lambda^+_k=\frac{N\!-\!k}{N}\cdot \frac{k}{N}\cdot \sum_{u_1,u_2\in\mathcal{U}} x(u_1,u_2)\cdot \rho(u_1,u_2).
\end{equation}
In this expression, $(N\!-\!k)/N$ is the probability that the randomly chosen learner is a resident, and $k/N$ is the probability that the role model is a mutant. The sum corresponds to the total probability that the learner adopts the role model's strategy over all possible payoffs $u_1$ and $u_2$ that the two player may have received in their respective last rounds. We use $x(u_1,u_2)$ to denote the probability that the randomly chosen resident obtained a payoff of $u_1$ in the last round of his respective game, and that the mutant obtained a payoff of $u_2$. Given that the payoffs are $u_1$ and $u_2$, the imitation probability is then given by $\rho(u_1,u_2)$, as specified by Eq.~(\ref{Eq:rho}). The probability that the respective payoffs of the players are given by $u_1$ and $u_2$ can be calculated as
\begin{equation}
\setlength{\arraycolsep}{1pt}
\begin{array}{llrl}
x(u_1,u_2)	 &=&\displaystyle \frac{1}{N\!-\!1}\cdot  &v_{u_1}(S_1,S_2)\cdot 1_{(u_1,u_2)\in \mathcal{U}^2_F}\\[0.5cm]
&+	
&\displaystyle \left(1\!-\!\frac{1}{N\!-\!1}\right)  
&\left[ \frac{k\!-\!1}{N\!-\!2}\frac{k\!-\!2}{N\!-\!3} v_{u_1}(S_1,S_2) v_{u_2}(S_2,S_2) + 
 \frac{k\!-\!1}{N\!-\!2}\frac{N\!-\!k\!-\!1}{N\!-\!3} v_{u_1}(S_1,S_2) v_{u_2}(S_2,S_1)\right.\\[0.5cm]
&&&\left. +\frac{N\!-\!k\!-\!1}{N\!-\!2}\frac{k\!-\!1}{N\!-\!3} v_{u_1}(S_1,S_1) v_{u_2}(S_2,S_2) + 
 \frac{N\!-\!k\!-\!1}{N\!-\!2}\frac{N\!-\!k\!-\!2}{N\!-\!3} v_{u_1}(S_1,S_1) v_{u_2}(S_2,S_1)\right].
\end{array}
\end{equation}
The first term on the right side corresponds to the case that the learner and the role model happened to be matched during the game stage, which happens with probability $1/(N\!-\!1)$. In that case, we note that only those payoff pairs can occur that are feasible in a direct interaction, $(u_1,u_2)\in \mathcal{U}^2_F:=\big\{ (R,R), (S,T), (T,S), (P,P) \big\}$, as represented by the respective indicator function. Otherwise, if the learner and the role model did not interact directly, we need to distinguish four different cases, depending on whether the learner was matched with a resident or a mutant, and depending on whether the role model was matched with a resident or a mutant.
